---
title: "Assignment 9"
subtitle: "Decision analysis"
author: "Aki Vehtari et al."
format:
  html:
    toc: true
    code-tools: true
    code-line-numbers: true
    number-sections: true
    mainfont: Georgia, serif
    page-layout: article
editor: source
filters:
  - includes/assignments.lua 
  - includes/include-code-files.lua 
---

# General information

**The maximum amount of points from this assignment is 3.**

We have prepared a **quarto template specific to this assignment ([html](template9.html), [qmd](https://avehtari.github.io/BDA_course_Aalto/assignments/template9.qmd), [pdf](template9.pdf))** to help you get started. 


:::{.aalto}
We recommend you use [jupyter.cs.aalto.fi](https://jupyter.cs.aalto.fi) or the [docker container](docker.html).
:::


:::{.hint}
**Reading instructions**:

- [**The reading instructions for BDA3 Chapter 9**](../BDA3_notes.html#ch6) (decision analysis).

{{< include includes/_grading_instructions.md >}}

{{< include includes/_cmdstanr.md >}}
:::

{{< include includes/_reporting_accuracy.md >}}

{{< include includes/_general_info.md >}}






This exercise is an example of a decision analysis (DA). In a broad context, this means optimizing over different decisions that lead to different outcomes that all have different utilities. In a Bayesian context, this means using posterior distributions to make decisions.

# Escaping from the chicken coop

You are an adult chicken living in an organic chicken commune, where life is great, if a bit boring. 
You have settled in comfortably, but you want something more for your offspring. 
Your traveling corvid friends tell you of places where chickens eat corn all day, 
get ferried around in [mobile chicken coops to see the world](https://www.chicken-trailer.de/en/chicken-trailer-en/chicken-trailer-330) or [get mental stimulation by being trained humanely to perform
tricks](https://en.wikipedia.org/wiki/Marian_Breland_Bailey#:~:text=Popular%20acts%20included,that%20played%20basketball.).
The chicken elders have gained access to the computers of your human caretakers
and have found the results of a complicated statistical analysis of the growth curves 
of your relatives. Because you are a chicken, you don't care about convergence diagnostics or priors.

**Your task is to maximizes the chance of escape for your offspring.**

::: {.subtask letter=a}
Compute and visualize the **expected chicken weight for days 1--40 per diet**, according to the
model provided in the template. Do the predictions look reasonable? Why/why not? 
:::
::: {.hint}
To sample a "new chicken" from the posterior, use [posterior_predict](https://rdrr.io/cran/brms/man/posterior_epred.brmsfit.html) with options `newdata=..., allow_new_levels=TRUE, sample_new_levels="gaussian"`, where you pass a dataframe as `newdata` which has a "new" chicken ID `Chick` and appropriate values for `Time` and `Diet`.
:::


:::{.subrubric}
* Does the plot look right and is it readable? It should be quite close to the reference plot below:
* Has it been recognized that the prediction time **...**?
:::

Your chicken elders have been meticulously collecting data on what kind of characteristics have allowed
previous chickens to escape. They have found out that both the age and the weight influence the (daily)
probability of escape for a chicken:

* If the chicken is too young, it is not yet mature enough to venture out into the world.
* If the chicken is too old, it will not try to escape anymore.
* If the chicken is small and has just the right size, it can try to squeze through a tiny crack in the fence.
* If the chicken is big enough, it is strong enough to try to fly over the fence.
* No matter the size, there is always a small residual probability that the chicken can escape.

Every day, chickens will try to escape if they are of the right age. 
Their daily escape probability $e(\text{day}, \text{weight})$ is implemented in the `daily_probability_of_escape(day, weight)` function.
The probability that a chicken with daily weights $w = (w_1,\dots,w_N)$ **has not escaped** after $i+1$ days can be computed as follows: 
$$
  f_{i+1} = f_i \, (1 - e(i, w_i))
$$
The `chickenwise_probability_of_escape(weights)` computes the probability that a chicken **has escaped** after `length(weights)` days.



:::: {.content-hidden when-format="pdf"}
::: {.callout-tip collapse=true}

###  Chickenwise probability of escape function

```{r}
bump <- function(x, loc=0, scale=1){
  xi = (x - loc) / scale
  ifelse(abs(xi) < 1, exp(-1/(1-xi^2)), 0.)
}
daily_probability_of_escape <- function(day, weight){
  # Expects a day and a weight and computes the daily probability of escape 
  bump(day, 30, 10) * (1e-2 + bump(weight, 200, 150)+bump(weight, 700, 150))
}
chickenwise_probability_of_escape <- function(weights){
  # Expects a vector of daily weights from day 1 to N and computes the probability of 
  # escape at the end of the time series
  prob_of_failure = 1
  for(day in 1:length(weights)){
    prob_of_failure = prob_of_failure * (1-daily_probability_of_escape(day, weights[day]))
  }
  return(1 - prob_of_failure)
}
```

:::
::::


```{r}
#| fig-cap: Daily probability of escape for a given day and weight
#| label: fig-prob
days = 1:40
weights = 1:900
heatmap_matrix = outer(days,weights,daily_probability_of_escape)
image(days, weights, heatmap_matrix, xlab="day", ylab="weight", main="Daily probability of escape")
```

::: {.subtask}
Compute and visualize the **distribution of the chickenwise probabilities of escape per diet**, 
according to the model provided in the template.
:::
::: {.hint}
You can reuse the predictions you created in the previous subtask. 
**Work with the draws to compute the chickenwise probability of escape first, and then take the expectation!** 
:::

:::{.subrubric}
* Does the plot look right and is it readable? It should be quite close to the reference plot below:
:::

::: {.subtask}
Compute the **expected probability of escape for each diet**. 
Why would it be wrong to compute the expected probability of escape by
applying the `chickenwise_probability_of_escape` function in the template to the 
*daily expected chicken weights per diet* computed in subtask 2.a? 
How does the correctly calculated value compare to the incorrectly calculated value?
Why is one higher than the other?
:::

:::{.subrubric}
* Do the results look correct and have they been presented in a readable way? They should be roughly the first column below for the correct calculation and the second column below for the wrong calculation: **...**
* Has it been explained that the expected probability of escape depends on **...**?
* Has it been explained that 

**...** is lower because **...**?

**...**

:::

{{< include includes/_overall_quality.md >}}
  