---
title: "Assignment 6"
author: "Aki Vehtari et al."
format:
  html:
    toc: true
    code-tools: true
    code-line-numbers: true
    number-sections: true
    mainfont: Georgia, serif
    page-layout: article
editor: source
filters:
  - includes/assignments.lua
  - includes/include-code-files.lua
---

# General information

**The maximum amount of points from this assignment is 6.**

We have prepared a **quarto template specific to this assignment ([html](template6.html), [qmd](https://avehtari.github.io/BDA_course_Aalto/assignments/template6.qmd), [pdf](template6.pdf))** to help you get started.


::: {.aalto}
We recommend you use [jupyter.cs.aalto.fi](https://jupyter.cs.aalto.fi) or the [docker container](docker.html).
:::

::: {.hint}
**Reading instructions:**

- [**The reading instructions for BDA3 Chapter 10**](../BDA3_notes.html#ch10).
- [**The reading instructions for BDA3 Chapter 11**](../BDA3_notes.html#ch11).

{{< include includes/_grading_instructions.md >}}

{{< include includes/_cmdstanr.md >}}
:::

{{< include includes/_reporting_accuracy.md >}}

{{< include includes/_general_info.md >}}


:::: {.content-hidden when-format="pdf"}
::: {.callout-tip collapse=true}

## Setup

JupyterHub has all the needed packages pre-installed.

The following installs and loads the `aaltobda` package:
```{r}
if(!require(aaltobda)){
    install.packages("aaltobda", repos = c("https://avehtari.github.io/BDA_course_Aalto/", getOption("repos")))
    library(aaltobda)
}
```
The following installs and loads the [`latex2exp` package](https://github.com/stefano-meschiari/latex2exp), which allows us to use LaTeX in plots:
```{r}
if(!require(latex2exp)){
    install.packages("latex2exp")
    library(latex2exp)
}
```
The following installs and loads the [`posterior` package](https://github.com/stan-dev/posterior) which imports the `rhat_basic()` function:
```{r}
if(!require(posterior)){
    install.packages("posterior")
    library(posterior)
}
```
The following installs and loads the [`ggplot2` package](https://ggplot2.tidyverse.org/), the [`bayesplot` package](https://mc-stan.org/bayesplot/index.html) and the [`dplyr` package](https://dplyr.tidyverse.org/)
```{r}
if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
if(!require(bayesplot)){
    install.packages("bayesplot")
    library(bayesplot)
}
if(!require(dplyr)){
    install.packages("dplyr")
    library(dplyr)
}
if(!require(tidyr)){
    install.packages("tidyr")
    library(tidyr)
}
# Some additional set-up to make plots legible
ggplot2::theme_set(theme_minimal(base_size = 14))
bayesplot::bayesplot_theme_set(theme_minimal(base_size = 14))
# register_knitr_engine()
```
The following installs and loads the [`cmdstanr` package](https://mc-stan.org/cmdstanr/) and tries to install `cmdstan`.
```{r}
if(!require(cmdstanr)){
    install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
    library(cmdstanr)
}
cmdstan_installed <- function(){
  res <- try(out <- cmdstanr::cmdstan_path(), silent = TRUE)
  !inherits(res, "try-error")
}
if(!cmdstan_installed()){
    install_cmdstan()
}
```

:::
::::


# Stan warm-up: linear model of BDA retention with Stan (2 points)

From 2018 to 2022, we have been keeping track of assignment submissions for the BDA course given the number of submissions for the 1st assignment. We will fit a simple linear model to answer two questions of interest:

- What is the trend of student retention as measured by assignment submissions?
- Given the submission rates for assignments 1--8, how many students will complete the final 9th assignment (and potentially pass the course)?

The author has given you the broken Stan code below, which they intend to encode the following linear model:
$$
\begin{aligned}
p(\alpha,\beta,\sigma) &= \mathrm{const.}
      & \text{(improper flat prior)}&\text{ and}\\
p(y|x,\alpha,\beta,\sigma) &= p_\mathrm{normal}(y|\alpha + \beta x, \sigma)
      & \text{(normal likelihood)} &\text{.}
\end{aligned}
$$
In both the statistical model above and in the Stan model below, $x \in \mathbb{R}^N$ and $y \in \mathbb{R}^N$ are vectors of the covariates / predictors (the assignment number) and vectors of the observation (proportions of students who have handed in the respective assignment). $\alpha \in \mathbb{R}$ is the unknown scalar intercept, $\beta \in \mathbb{R}$ is the unknown scalar slope and $\sigma \in \mathbb{R}_{>0}$ is the unknown scalar observation standard deviation. The statistical model further implies
$$
p(y_\mathrm{pred.}|x_\mathrm{pred.},\alpha,\beta,\sigma) = p_\mathrm{normal}(y_\mathrm{pred.}|\alpha + \beta x_\mathrm{pred.}, \sigma)
$$
as the predictive distribution for a new observation $y_\mathrm{pred.}$ at a given new covariate value $x_\mathrm{pred.}$.

You can download [the broken stan file from github](./additional_files/assignment6/linear_model.stan).

```{.stan}
data { #<1>
    // number of data points
    int<lower=0> N;
    // covariate / predictor
    vector[N] x;
    // observations
    vector[N] y;
    // number of covariate values to make predictions at
    int<lower=0> no_predictions;
    // covariate values to make predictions at
    vector[no_predictions] x_predictions;
} #<1>
parameters { #<2>
    // intercept
    real alpha;
    // slope
    real beta;
    // the standard deviation should be constrained to be positive
    real<upper=0> sigma;
} #<2>
transformed parameters { #<3>
    // deterministic transformation of parameters and data
    vector[N] mu = alpha + beta * x // linear model
} #<3>
model { #<4>
    // observation model / likelihood
    y ~ normal(mu, sigma);
} #<4>
generated quantities { #<5>
    // compute the means for the covariate values at which to make predictions
    vector[no_predictions] mu_pred = alpha + beta * x_predictions;
    // sample from the predictive distribution, a normal(mu_pred, sigma).
    array[no_predictions] real y_pred = normal_rng(mu, sigma);
} #<5>
```

1. This is [Stan's data block](https://mc-stan.org/docs/reference-manual/program-block-data.html):
"*The `data` block is for the declaration of variables that are read in as data. [...] Each variable’s value is validated against its declaration as it is read. For example, if a variable sigma is declared as `real<lower=0>`, then trying to assign it a negative value will raise an error. As a result, data type errors will be caught as early as possible. Similarly, attempts to provide data of the wrong size for a compound data structure will also raise an error.*" For more information, follow the link.

2. This is [Stan's parameters block](https://mc-stan.org/docs/reference-manual/program-block-parameters.html):
"*The variables declared in the `parameters` program block correspond directly to the variables being sampled by Stan’s samplers (HMC and NUTS). From a user’s perspective, the parameters in the program block are the parameters being sampled by Stan.*" For more information, follow the link.

3. This is [Stan's transformed parameters block](https://mc-stan.org/docs/reference-manual/program-block-transformed-parameters.html):
"*The `transformed parameters` program block consists of optional variable declarations followed by statements. After the statements are executed, the constraints on the transformed parameters are validated. Any variable declared as a transformed parameter is part of the output produced for draws.*" For more information, follow the link.

4. This is [Stan's model block](https://mc-stan.org/docs/reference-manual/program-block-model.html):
"*The `model` program block consists of optional variable declarations followed by statements. The variables in the `model` block are local variables and are not written as part of the output. [...] The statements in the `model` block typically define the model. This is the block in which probability (sampling notation) statements are allowed.*" For more information, follow the link.

5. This is [Stan's generated quantities block](https://mc-stan.org/docs/reference-manual/program-block-generated-quantities.html):
"*The `generated quantities` program block is rather different than the other blocks. Nothing in the `generated quantities` block affects the sampled parameter values. The block is executed only after a sample has been generated.*" For more information, follow the link.

::: {.hint}
A normal linear model is actually **not** the best model to use for this type of data, but we will use it here to illustrate the first step in building up to more appropriate, complicated models.
:::

::: {.subtask}
<!-- ### Find mistakes in the code -->
Find the ***three mistakes*** in the code and fix them. Report the original mistakes and your fixes clearly in your report. Include the **full** corrected Stan code in your report. Verify that sampling was successful.
:::

::: {.hint}
You may find some of the mistakes in the code using Stan syntax checker. If you copy the Stan code to a file ending `.stan` and open it in RStudio (you can also choose from RStudio menu File$\rightarrow$New File$\rightarrow$Stan file to create a new Stan file), the editor will show you some syntax errors. More syntax errors might be detected by clicking `Check' in the bar just above the Stan file in the RStudio editor. Note that some of the errors in the presented Stan code may not be syntax errors.
:::



<!-- ### Interpret the results -->
The author runs the corrected Stan file using the following R code and plots the returned MCMC sample. Read through the code below to understand what is being plotted.


:::: {.content-hidden when-format="pdf"}
::: {.callout-tip collapse=true}

## Data preparation and sampling from the posterior

**Data assembly happens here**:
```{r}
#| warning: false
# These are our observations y: the proportion of students handing in each assignment (1-8),
# sorted by year (row-wise) and assignment (column-wise).
# While the code suggest a matrix structure,
# the result will actually be a vector of length N = no_years * no_assignments
propstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123)/176,
                c(242, 212, 184, 177, 174, 172, 163, 156)/242,
                c(332, 310, 278, 258, 243, 242, 226, 224)/332,
                c(301, 269, 231, 232, 217, 208, 193, 191)/301,
                c(245, 240, 228, 217, 206, 199, 191, 182)/245)
# These are our predictors x: for each observation, the corresponding assignment number.
assignment <- rep(1:8, 5)
# These are in some sense our test data: the proportion of students handing in the last assignment (9),
# sorted by year.
# Usually, we would not want to split our data like that and instead
# use e.g. Leave-One-Out Cross-Validation (LOO-CV, see e.g. http://mc-stan.org/loo/index.html)
# to evaluate model performance.
propstudents9 = c(121/176, 153/242, 218/332, 190/301, 175/245)
# The total number of assignments
no_assignments = 9
# The assignment numbers for which we want to generate predictions
x_predictions = 1:no_assignments
# (Cmd)Stan(R) expects the data to be passed in the below format:
model_data = list(N=length(assignment),
                 x=assignment,
                 y=propstudents,
                 no_predictions=no_assignments,
                 x_predictions=x_predictions)
```
**Sampling from the posterior distribution happens here**:
```{r}
#| warning: false
# This reads the file at the specified path and tries to compile it.
# If it fails, an error is thrown.
retention_model = cmdstan_model("./additional_files/assignment6_linear_model.stan")
# This "out <- capture.output(...)" construction suppresses output from cmdstanr
# See also https://github.com/stan-dev/cmdstanr/issues/646
out <- capture.output(
    # Sampling from the posterior distribution happens here:
    fit <- retention_model$sample(data=model_data, refresh=0, show_messages=FALSE)
)
```
**Draws postprocessing happens here**:
```{r}

# This extracts the draws from the sampling result as a data.frame.
draws_df = fit$draws(format="draws_df")

# This does some data/draws wrangling to compute the 5, 50 and 95 percentiles of
# the mean at the specified covariate values (x_predictions).
# It can be instructive to play around with each of the data processing steps
# to find out what each step does, e.g. by removing parts from the back like "|>  gather(pct,y,-x)"
# and printing the resulting data.frame.
mu_quantiles_df = draws_df |>
      subset_draws(variable = c("mu_pred")) |>
      summarise_draws(~quantile2(.x, probs = c(0.05, .5, 0.95))) |>
      mutate(x = 1:9) |>
      pivot_longer(c(q5, q50, q95), names_to = c("pct"))
# Same as above, but for the predictions.
y_quantiles_df = draws_df |>
      subset_draws(variable = c("y_pred")) |>
      summarise_draws(~quantile2(.x, probs = c(0.05, .5, 0.95))) |>
      mutate(x = 1:9) |>
      pivot_longer(c(q5, q50, q95), names_to = c("pct"))
```

:::
::::

::: {.both}
**Plotting happens here**:
```{r}
#| label: fig-posterior
#| fig-cap: Describe me in your submission!
ggplot() +
  # scatter plot of the training data:
  geom_point(
    aes(x, y, color=assignment),
    data=data.frame(x=assignment, y=propstudents, assignment="1-8")
) +
  # scatter plot of the test data:
  geom_point(
    aes(x, y, color=assignment),
    data=data.frame(x=no_assignments, y=propstudents9, assignment="9")
) +
  # you have to tell us what this plots:
  geom_line(aes(x,y=value,linetype=pct), data=mu_quantiles_df, color='grey', linewidth=1.5) +
  # you have to tell us what this plots:
  geom_line(aes(x,y=value,linetype=pct), data=y_quantiles_df, color='red') +
  # adding xticks for each assignment:
  scale_x_continuous(breaks=1:no_assignments) +
  # adding labels to the plot:
  labs(y="assignment submission %", x="assignment number") +
  # specifying that line types repeat:
  scale_linetype_manual(values=c(2,1,2)) +
  # Specify colours of the observations:
  scale_colour_manual(values = c("1-8"="black", "9"="blue")) +
  # remove the legend for the linetypes:
  guides(linetype="none")
```
:::

:::: {.content-hidden when-format="pdf"}
::: {.callout-tip collapse=true}

## Quick check for sampling convergence

If your model is correctly implemented, sampling from the posterior distribution should have been successful.
You can check whether Stan thinks that sampling succeeded by inspecting the output of the below command,
which you should be able to interpret with a little help from the [CmdStan User's Guide](https://mc-stan.org/docs/cmdstan-guide/diagnose.html).
```{r}
fit$cmdstan_diagnose()
```

:::
::::


Based on the above plot, answer the following questions:

::: {.subtask}
- What is the solid red line plotting? What are the dashed red lines? How and why are these different from the corresponding grey lines?
- What is the general trend of student retention as measured by assignment submissions?
- Given a model fitted to the submission data for assignments 1-8, does it do a good job predicting the proportion of students who submit the final 9th assignment?
- Name one different modeling choice you could make to improve the prediction.
:::


::: {.rubric}
* Is the source code included?
    * No
    * Yes
* Is the full resulting modified Stan model code presented in the report?
    * No
    * Yes, but partially
    * Yes, with a few mistakes
    * Yes, and it is correct
* Has the sampling success been verified/summarized (e.g. by inspecting and summarizing the output of [`CmdStan`'s `diagnose` method](https://mc-stan.org/docs/cmdstan-guide/diagnose.html)).
    * No
    * Yes, but partially
    * Yes, and it is correctly verified
* Fix #1: Is there a fix for line **...**.
    * It has not been discussed, that this line should be fixed.
    * It has been discussed, that this line should be fixed, but there is no fix presented for it or the fix is clearly wrong.
    * There is a fix presented for this line, that clearly solves the problem.
* Fix #2: Is there a fix for line **...**.
    * It has not been discussed, that this line should be fixed.
    * It has been discussed, that this line should be fixed, but there is no fix presented for it or the fix is clearly wrong.
    * There is a fix presented for this line, that clearly solves the problem.
* Fix #3: Is there a fix for line **...**.
    * It has not been discussed, that this line should be fixed.
    * It has been discussed, that this line should be fixed, but there is no fix presented for it or the fix is clearly wrong.
    * There is a fix presented for this line, that clearly solves the problem.
* Have the red lines been correctly described (**...**)?
* Have the grey lines been correctly described (**...**)?
* Has the difference between the red and grey lines been explained (**...**)?
* Has the student retention trend been described (**...**)
* Has the predictive performance for the held out data been discussed and assessed satisfactorily (**...**)?
    * No
    * Somewhat
    * Yes
* Has at least one way to improve the model been mentioned (E.g. **...** or **...**)?
:::

# Generalized linear model: Bioassay with Stan (4 points)

Replicate the computations for the bioassay example of section 3.7
(BDA3) using Stan.

::: {.subtask}
Write down the model for the bioassay data in Stan syntax. For
instructions in reporting your implementation, you can refer to
parts 2 c) - g) in Assignment 5. Use the Gaussian prior as in Assignment 4 and 5, that is
$$
\begin{aligned}
    \begin{bmatrix}
    \alpha \\ \beta
    \end{bmatrix}
    \sim
    \text{N} \left( \mu_0,  \Sigma_0 \right), \qquad
    \text{where} \quad
     \mu_0 = \begin{bmatrix} 0 \\ 10 \end{bmatrix} \quad \text{and} \quad
     \Sigma_0 = \begin{bmatrix} 2^2 & 12 \\ 12 & 10^2 \end{bmatrix}.
\end{aligned}
$$
:::

::: {.hint}
You will need Stan functions `multi_normal` and
`binomial_logit` for implementing the prior and observation model,
respectively. In Stan code, it is easiest to declare a variable (say
`theta`) which is a two-element vector so that the first value
denotes $\alpha$ and latter one $\beta$. This is because the
`multi_normal` function that you need for implementing the prior
requires a vector as an input.
:::


::: {.subtask}
Use $\widehat{R}$ for convergence analysis. You can either use Eq.
(11.4) in BDA3 or the later version that can be found in [**a recent
article**](https://arxiv.org/abs/1903.08008). You should specify
which $\widehat{R}$ you used. In R the best choice is to use
function `rhat_basic()` or `rhat()` from the `posterior` package
(see `?posterior::rhat_basic`). To check $\widehat{R}$ and other
diagnostics, you can also call `fit$summary()`, where `fit` is the
fit object returned by Stan's sampling function. Report the
$\widehat{R}$ values both for $\alpha$ and $\beta$ and discuss the
convergence of the chains. **Briefly explain in your own words how
to interpret the obtained** $\widehat{R}$ **values**.
:::



::: {.subtask}
Plot the draws for $\alpha$ and $\beta$ (scatter plot) and include
this plot in your report. You can compare the results to Figure 3.3b
in BDA3 to verify that your code gives sensible results. Notice
though that the results in Figure 3.3b are generated from posterior
with a uniform prior, so even when your algorithm works perfectly,
the results will look slightly different (although fairly similar).
:::



::: {.subtask}
To develop the course and provide feedback to Stan developers, we
collect information on which Stan setup you used and whether you had
any problems in setting it up or using it. Please report,

-   Operating system (Linux, Mac, Windows) [or jupyter.cs.aalto.fi]{.aalto}?

-   Programming environment used: R or Python?

-   Interface used: RStan, CmdStanR, PyStan, or CmdStanPy?

-   Did you have installation or compilation problems? [Did you try first installing locally, but switched to
    jupyter.cs.aalto.fi?]{.aalto}

-   In addition of these you can write what other things you found
    out difficult (or even frustrating) when making this assignment
    with Stan.
:::




::: {.rubric}
* Is the Stan model code included?
    * No
    * Yes
* Does the implemented Stan-model seem to be working?
    * No implementation
    * Model implemented but results not visualized/reported
    * Model implemented, but the results seem weird
    * Model seems to work correctly
* Are the R_hat-values reported (potential scale reduction factor, Eq. (11.4) in the BDA3)?
    * No
    * Yes, but only for alpha or beta
    * Yes, single values both for alpha and beta
* Is the interpretation of R_hat values correct (**...**)?
    * No interpretation or discussion about the R_hat-values, or conclusions clearly wrong
    * Somewhat correct
    * Interpretation correct
* Does the report contain a scatter plot about the draws? Do the results look reasonable, that is, roughly
like in the Figure below **...**?
    * No plot included
    * Plot included, but the results do not look like in the figure above
    * Plot included, and the results look roughly like in the figure above
* Does the report contain description of Stan setup used and whether there were any problems in setting
it up or using it?
    * No
    * Yes
* Even if the Stan model code is correct, there might be ways to give improve the layout or write the
model in more elegant ways. This optional feedback box can be used to give additional suggestions for
better Stan code.
:::

{{< include includes/_overall_quality.md >}}
