---
title: "Assignment 1 - Bayesian data analysis"
author: anonymous # <-- hand in anonymously
format: 
  html:
    toc: true
    code-tools: true
    code-line-numbers: true
    standalone: true
    self-contained: true
    embed-resources: true  
    number-sections: true
    mainfont: Georgia, serif
    fontsize: 1.2em
  pdf:  
    documentclass: article
    number-sections: true
reference-location: margin
citation-location: margin 
comments:
  hypothesis: true
editor: source
---


# General information    

:::: {.content-hidden when-format="pdf"}
::: {.callout-warning}
   
## Setup 

*This block will only be visible in your HTML output, but will be hidden when rendering to PDF for the submission.*



This is the template for [assignment 1](assignment1.html). You can download the [qmd-file](https://raw.githubusercontent.com/avehtari/BDA_course_Aalto/gh-pages/assignments/template1.qmd) or copy the code from this rendered document after clicking on `</> Code` in the top right corner.

**Please replace the instructions in this template by your own text, explaining what you are doing in each exercise.** 

The following will set-up `markmyassignment` to check your functions at the end of the notebook:
 
```{r} 
library(markmyassignment) 
assignment_path = paste("https://github.com/avehtari/BDA_course_Aalto/",
"blob/master/assignments/tests/assignment1.yml", sep="")
set_assignment(assignment_path)   
```       

:::
::::


# Basic probability theory notation and terms  

 
**Write your answers here!** 
   


# Basic computer skills

**Do some setup here. Explain shortly what you do.**
```{r}
# Do some setup:
distribution_mean = .2
distribution_variance = .01

# You have to compute the parameters below from the given mean and variance
# distribution_alpha = ...
# distribution_beta = ...
```


## (a)


**Plot the PDF here. Explain shortly what you do.**
```{r}
# Useful functions: seq(), plot() and dbeta()
```


## (b)


**Draw samples and plot the histogram here. Explain shortly what you do.**
```{r}
# Useful functions: rbeta() and hist()
```


## (c)


**Compute the sample mean and variance here. Explain shortly what you do.**
```{r}
# Useful functions: mean() and var()
```


## (d)


**Compute the central interval here. Explain shortly what you do.**
```{r}
# Useful functions: quantile()
```


# Bayes' theorem 1   



**Compute the quantities needed to justify your recommendation here. Explain shortly what you do.**
```{r}
# You can do the computation with pen and paper or in R. 
# Either way, you have to explain why you compute what you compute.
```


# Bayes' theorem 2

**You will need to change the numbers to the numbers in the exercise.**
```{r}
boxes <- matrix(c(2,2,1,5,5,1), ncol = 2,
    dimnames = list(c("A", "B", "C"), c("red", "white")))
```


## (a)


**Keep the below name and format for the function to work with `markmyassignment`:**
```{r}
p_red <- function(boxes) {
    # Do computation here, and return as below.
    # This is a test case, you will need to change 
    # the numbers to the numbers in the exercise. 
    0.3928571
}
```


## (b)


**Keep the below name and format for the function to work with `markmyassignment`:**
```{r}
p_box <- function(boxes) {
    # Do computation here, and return as below.
    # This is a test case, you will need to change 
    # the numbers to the numbers in the exercise. 
    c(0.29090909,0.07272727,0.63636364)
}
```


# Bayes' theorem 3 



**You will need to change the numbers to the numbers in the exercise.**
```{r}
fraternal_prob = 1/125
identical_prob = 1/300
```
**Keep the below name and format for the function to work with `markmyassignment`:**
```{r}
p_identical_twin <- function(fraternal_prob, identical_prob) {
    # Do computation here, and return as below.
    # This is a test case, you will need to change 
    # the numbers to the numbers in the exercise. 
    0.4545455
}
```



:::: {.content-hidden when-format="pdf"}
::: {.callout-warning}

## markmyassignment

*This block will only be visible in your HTML output, but will be hidden when rendering to PDF for the submission.*


The following will check the functions for which `markmyassignment` has been set up:
 
```{r} 
mark_my_assignment()  
```      

:::
::::

