\documentclass[a4paper,11pt]{article}

\usepackage[pdftex]{graphicx}
%\usepackage{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage[T1,mtbold,lucidacal,mtplusscr,subscriptcorrection]{mathtime}
\usepackage{times}
\usepackage{amsmath}
\usepackage[hyphens]{url}
\usepackage{enumerate}
\usepackage{parskip}
\usepackage[colorlinks,urlcolor=black]{hyperref}
\usepackage{microtype}

%\usepackage[dvips,bookmarks=false]{hyperref}
% \hypersetup{%
%   bookmarksopen=true,
%   bookmarksnumbered=true,
%   pdftitle={S-114.2601 Bayesilaisen mallintamisen perusteet},
%   pdfsubject={Kommentteja},
%   pdfauthor={Aki Vehtari},
%   pdfkeywords={bayesilainen todennäköisyysteoria, bayesilainen
%     päättely, bayesilaiset mallit, mallien analysointi,
%     laskennalliset menetelmät, Markov-ketju Monte Carlo},
%   pdfstartview={FitH -32768}
% }


% if not draft, smaller printable area makes the paper more readable
\topmargin -4mm
\oddsidemargin 0mm
\textheight 225mm
\textwidth 150mm

%\parskip=\baselineskip

\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\Sd}{Sd}
\DeclareMathOperator{\sd}{sd}
\DeclareMathOperator{\Bin}{Bin}
\DeclareMathOperator{\Beta}{Beta}
\DeclareMathOperator{\Poisson}{Poisson}
\DeclareMathOperator{\betacdf}{betacdf}
\DeclareMathOperator{\Invchi2}{Inv-\chi^2}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\N}{N}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\trace}{trace}

\pagestyle{empty}

\begin{document}
\thispagestyle{empty}

\section*{Bayesian data analysis -- exercise 9}
This assignment is related to Chapter 9.

The maximum amount of points from this assignment is 3. In addition to the correctness of the answers, the overall quality and clearness of the report is evaluated.

Report all results to a single, {\bf anonymous} *.pdf -file and return it to \href{peergrade.io}{peergrade.io}. Include also source code to the report (either as an attachment or as a part of the answer). By anonymity it is meant that the report should not contain your name or student number.

\vspace{1cm}

\subsection*{Decision analysis for the factory data (3p)}

This exercise is an example of decision analysis (DA). In broad context, this means optimizing over different decisions that lead to different outcomes that all have different utilities. In Bayesian context, this means using posterior distributions to make decisions.  

In this exercise, you work as a data analyst in the company that owns the six machines that have produced the data in \texttt{factory.txt}. Your task is to decide whether or not to buy a new (7th) machine for the company. The decision should be based on our best knowledge about the machines.

The following is known about the production process: 
\begin{itemize} 
\item The given data contains quality measurements of single products from the six machines that are ordered from the same seller.
\item Making each product costs 100 euros.
\item Customers pay 200 euros for each product.
\item If the quality of the product is below 85, the product cannot be sold 
\item All the products that have sufficient quality are sold.
\item The cost of acquiring the new machine is low compared to its expected usage time so this cost is not taken into account in this analysis.
\item The only thing the company owner cares  about is money. Thus, as a utility function, use the profit of a new product from a machine.
\end{itemize}

Your task is the following:
\begin{itemize}
\item For each of the six machines, compute and report the expected utility of the products of that machine.
\item Rank the machines based on the expected utilities and explain briefly what these values tell about the quality of these machines. E.g. Tell which machines are profitable and which are not (if any).
\item Compute and report the expected utility of the products of a new (7th) machine.
\item Based on your analysis, discuss briefly whether the company owner should buy a new (7th) machine.
\item As usual, remember to include the source code (for Stan and Python/R/Matlab)!
\end{itemize}

Hints and further information:
\begin{itemize}
\item As noticed in the previous assignment, the hierarchical model fits best to the dataset, so use it to compute the utilities.
%\item Don't compute the expected utility of the machines in Stan. Instead, take the posterior predictive samples given by Stan and use those in Python/R/Matlab. Computing expected utilities in Stan might result to 
\item Assume that the quality of the products of different machines have a common variance.
\item Use uniform priors for hyper-parameters as in previous assignment.
\end{itemize}

The assumptions for the hierarchical model are exactly the same as in the previous assignments! If you did things correctly then, solving the assignment does not require changing the Stan model. 

\end{document}
