---
title: "BDA - Assignment X"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
---

This is a template with format instructions for Assignments in the Bayesian Data Analysis course at Aalto University. R markdown is a convinient way of writing by combining text and R code using markdown syntax. To create your assignemtn, remove the formating instructions and use this file as a template.

More information on how to use markdown, see [this](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet) and more information on R markdown can be found [here](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

**Note** The report should be anonymous and submitted to [peergrade.io](peergrade.io) as `assignmentX.pdf`. If you have problem with creating a PDF file directly from R markdown, start by creating an HTML file and the just printe the HTML to a PDF. 


```{r setup, include=FALSE}
# This chunk just sets echo = TRUE as default (i.e. print all code)
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

All exercises in the assignment should start with header 1 fully specifying that it is exercise X.

## a)

Subtasks in each assignments should be numbered and use header 2 (two ##). 

We can easy add R code as chunks in the following way

```{r cars}
5 + 5
```

Include plots

```{r pressure}
plot(pressure)
```



