---
title: "Assignment 2, 2023"
author: "Aki Vehtari et al."
format:
  html:
    toc: true
    code-tools: true
    code-line-numbers: true
    number-sections: true
    mainfont: Georgia, serif
  pdf:
    geometry:
    - left=1cm,top=1cm,bottom=1cm,right=7cm
    number-sections: true
    code-annotations: none
editor: source
---


# General information

**This is for BDA 2023**

This assignment is related to Lecture 2 and BDA3 Chapters 1 and 2.
You may find an additional discussion
about choosing priors in a [**blog post by Andrew
Gelman**](http://andrewgelman.com/2017/10/04/worry-rigged-priors/).

**The maximum amount of points from this assignment is 3.**

We prepared a **quarto template specific to this assignment ([html](template2.html), [qmd](https://avehtari.github.io/BDA_course_Aalto/assignments/template2.qmd), [pdf](template2.pdf))** to help you get started.

::: {.callout-tip collapse=false}
**Reading instructions**:

- [**The reading instructions for BDA3 Chapter 1**](../BDA3_notes.html#ch1).
- [**The reading instructions for BDA3 Chapter 2**](../BDA3_notes.html#ch2).

{{< include includes/_grading_instructions.md >}}
:::


{{< include includes/_general_info.md >}}

::: {.callout-warning collapse=false}
## Setup

This is the template for [assignment 2](assignment2.html). You can download the [qmd-file](https://avehtari.github.io/BDA_course_Aalto/assignments/template2.qmd) or copy the code from this rendered document after clicking on `</> Code` in the top right corner.

**Please replace the instructions in this template by your own text, explaining what you are doing in each exercise.**

The following will set-up `markmyassignment` to check your functions at the end of the notebook:

```{.r}
library(markmyassignment)
assignment_path = paste("https://github.com/avehtari/BDA_course_Aalto/",
"blob/master/assignments/tests/assignment2.yml", sep="")
set_assignment(assignment_path)
```

The following installs the `aaltobda` package:
```{.r}
#| cache: true
# Caching should be fine here
install.packages("aaltobda", repos = c("https://avehtari.github.io/BDA_course_Aalto/", getOption("repos")))
```
:::

# Inference for binomial proportion

Algae status is monitored in 274 sites at Finnish lakes and rivers. The
observations for the 2008 algae status at each site are presented in
the dataset `algae` in the `aaltobda` package ('0': no algae, '1': algae present).


::: {.content-visible when-profile="public"}
Loading the library and the data.
```{r}
library(aaltobda)
data("algae")
# The data are now stored in the variable `algae`.
# These are the values for the prior required in the assignment
prior_alpha = 2
prior_beta = 10
```

The below data is **only for the tests**, you need to change to the
full data `algae` when reporting your results.
```{.r}
algae_test <- c(0, 1, 1, 0, 0, 0)
```
:::

Let $\pi$ be the probability of a monitoring site having detectable
blue-green algae levels and $y$ the observations in `algae`. Use a
binomial model for the observations $y$ and a $Beta(2,10)$ prior for
binomial model parameter $\pi$ to formulate a Bayesian model. Here it is
not necessary to derive the posterior distribution for $\pi$ as it has
already been done in the book and it suffices to refer to that
derivation. Also, it is not necessary to write out the distributions; it
is sufficient to use label-parameter format, e.g. $Beta(\alpha,\beta)$.

Your task is to perform Bayesian inference for a binomial model and answer
questions based on it:

::: {.callout-warning icon=false title="Subtask 2.a)"}
Formulate

1. the likelihood $p(y|\pi)$ as a function of $\pi$,
2. the prior $p(\pi)$, and
3. the resulting posterior $p(\pi|y)$.

Report the posterior in the format $Beta(\alpha,\beta)$, where you
replace $\alpha$ and $\beta$ with the correct numerical values.
:::

::: {.callout-tip collapse=false}
With a conjugate prior, a closed-form posterior has Beta form
(see equations in BDA3 and in the slides).
:::

::: {.content-visible when-profile="public"}
Write the likelihood, the prior and the posterior here!
```{r}
# These are not the actual values for the posterior!
# You will have to compute those from the data!
posterior_alpha = 2
posterior_beta = 10
```

You can do [string interpolation](https://en.wikipedia.org/wiki/String_interpolation) using [R inline code execution in quarto](https://quarto.org/docs/computations/execution-options.html#knitr) as such:

$\alpha_\text{prior}$ is **`r prior_alpha`** and $\beta_\text{prior}$ is **`r prior_beta`**.
Or string interpolation within math: $Beta(`r prior_alpha`,`r prior_beta`)$
:::

::: {.callout-warning icon=false title="Subtask 2.b)"}
What can you say about the value of the unknown $\pi$ according to
the observations and your prior knowledge? Summarize your results
with a point estimate (i.e. $E(\pi|y)$) and a 90% posterior
interval.
:::

::: {.callout-tip collapse=false}
Posterior intervals are also called credible
intervals and are different from confidence intervals.
:::

::: {.content-visible when-profile="public"}
Keep the below name and format for the functions to work with `markmyassignment`:
```{.r}
# Useful function: qbeta()

beta_point_est <- function(prior_alpha, prior_beta, data) {
    # Do computation here, and return as below.
    # This is the correct return value for the test data provided above,
    # combined with the prior provided above.
    0.2222222
}
beta_interval <- function(prior_alpha, prior_beta, data, prob=0.9) {
    # Do computation here, and return as below.
    # This is the correct return value for the test data provided above,
    # combined with the prior provided above.
    c(0.0846451, 0.3956414)
}
```
:::

::: {.callout-warning icon=false title="Subtask 2.c)"}
What is the probability that the proportion of monitoring sites with
detectable algae levels $\pi$ is smaller than $\pi_0=0.2$ that is
known from historical records?
:::
::: {.content-visible when-profile="public"}
Keep the below name and format for the function to work with `markmyassignment`:
```{.r}
# Useful function: pbeta()

beta_low <- function(prior_alpha, prior_beta, data, pi_0=0.2) {
    # Do computation here, and return as below.
    # This is the correct return value for the test data provided above,
    # combined with the correct prior.
    0.4511238
}
```
:::

::: {.callout-warning icon=false title="Subtask 2.d)"}
What assumptions are required in order to use this kind of a model
with this type of data?
:::

::: {.callout-tip collapse=false}
No need to discuss exchangeability yet, as
it is discussed in more detail in BDA3 Chapter 5 and Lecture 7.
:::

::: {.content-visible when-profile="public"}
Write your answer here!
:::

::: {.callout-warning icon=false title="Subtask 2.e)"}
Make prior sensitivity analysis by testing a couple of different
reasonable priors and plot the different posteriors. Summarize the
results by one or two sentences.
:::

::: {.content-visible when-profile="public"}
Plot the PDFs here. Explain shortly what you do.
```{.r}
# Useful function: dbeta()
```
:::


::: {.callout-note icon=false title="Rubric" weight=85}

* Is source code included?
* Are the prior, likelihood and posterior forms in a) reported (derivation of posterior not necessary)?
    * No
    * Some missing
    * Yes
* Is the reported resulting posterior correct ?
  * It is not reported, that the posterior distribution is a  distribution.
  * It is reported, that the posterior distribution is , but the numerical values for the
  parameters are incorrect
  * It is reported, that the posterior distribution is , and the numerical values for the
parameters are correct.
* In part b), is there at least one point estimate reported. Sample based estimates are also OK. Points
should be given if the method is right, even if the result is wrong due to a wrong posterior distribution
being used. With the right posterior, mean, median, and mode are all approximately .
* For the b) part, is the 90% posterior interval reported? Sample based estimate is also OK. Points
should be given if the method is right, even if the result is wrong because the posterior was wrong in
the first place. If the posterior was right, the 90% posterior interval is roughly .
* For the c) part, is the posterior probability Pr(π<0.2|y) reported? Points should be given if the method
is right, even if the result is wrong because the posterior was wrong. If the posterior was right, the
result should be approximately .
* For the d) part, does the report discuss [somehow the assumption of independence given the shared
success probability π (i.i.d. or independent and exchangeable trials is also OK)?]{.content-hidden when-profile="public"}
    * No
    * No, but other reasonable assumptions are discussed
    * Yes, but not quite right or some missing
    * Yes
* For the e) part, is there some comparison and discussion of results obtained with alternative prior
parameters?
    * No
    * Yes, but the results and conclusions are clearly wrong
    * Yes

:::

::: {.callout-warning collapse=false}
## markmyassignment
The following will check the functions for which `markmyassignment` has been set up:

```{.r}
mark_my_assignment()
```
:::

{{< include includes/_overall_quality.md >}}
